Genetic interactions in complex traits

### 2013-04-16 presentation/20120612_combio_seminar/combio_2012.tex
directory restructuring

\documentclass{beamer}
%\usepackage{colortbl}
\usepackage[latin1]{inputenc}
\usepackage{comment}
%\usetheme{Warsaw}
\usetheme{Frankfurt}
%\usecolortheme{dove}
\usefonttheme{serif}

\usepackage[absolute,overlay]{textpos} 
\usepackage{booktabs}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{multirow}
%\usepackage[table]{xcolor}
%\definecolor{tableShade}{HTML}{F1F5FA}
\usepackage[normalem]{ulem}

\title[Epistasis in GWAS]{2D or not 2D, that is the question\\
{\footnotesize Should we be looking for epistasis in GWAS?}}
\author{Gibran Hemani, Sara Knott, Joseph Powell, Naomi Wray, Peter Visscher, Chris Haley}
\institute{The Roslin Institute, University of Edinburgh \\
\vspace{.3cm}
Diamantina Institute \\and \\
Queensland Brain Institute,\\
University of Queensland}
\date{}
\begin{document}

@@

\begin{frame}{}
\begin{itemize}
\item \textbf{Simulations:} An evolutionary perspective on genetic variation
\item \textbf{Computation:} Making epistatic searches possible
\item \textbf{Thresholds:} What should we call `significant'?
\item \textbf{Data:} What do we find empirically?
\end{itemize}
\end{frame}

\section{Evolutionary simulations}
\subsection{}

@@

\begin{frame}{Epistasis and additive variance}
\begin{center}
\includegraphics[height=7cm]{gp_va}
\end{center}
\end{frame}

\begin{frame}{}
\textbf{Hypothesis:}\\ Epistasis can maintain additive variance better than purely additive effects
\end{frame}

\begin{frame}{Epistasis maintains genetic variation}
\begin{center}
\includegraphics[width=11cm]{maps_freqs}
\end{center}
\end{frame}

\begin{frame}{Most of the variance is non-additive}
\begin{center}
\includegraphics[width=8cm]{sup_propadditive_det.png}
\end{center}
\end{frame}

@@

\begin{frame}{With high LD, 2D searches are most powerful}
\begin{columns}[c]
\column{.7\textwidth}
\includegraphics[width=8cm]{sup_heritbars_sim.png}
\column{.3\textwidth}
When LD between causal variants and observed SNPs is high, 2D scan is most powerful for 46/54 non-additive patterns
\end{columns}
\end{frame}

@@

\begin{frame}{The problem}
Number of interactions: \\
\begin{equation}
n(n-1) / 2 \nonumber
\end{equation}
500000 SNPs $\rightarrow$ 125 billion interactions \\
\begin{itemize}
\item For 1000 individuals, PLINK performs 5000 tests per second $\rightarrow$ 3 years
\item FastEpistasis performs 36000 tests per second $\rightarrow$ 5 months
\item Optimised C code 125000 tests per second $\rightarrow$ 6 weeks
\item Parallelised on 8-core CPU $\rightarrow$ 5 days
\end{itemize}
\end{frame}

@@

\begin{frame}{Graphics Cards}
\begin{center}
\includegraphics[width=10cm]{gtx580}
\end{center}
\end{frame}

@@

\begin{frame}{Other optimised epistasis software}
\begin{itemize}
\item BiForce
\item GLIDE
\item GBOOST
\item TEAM
\item MDR-GPU
\item EPIBLASTER
\item SHEsisEPI
\item ...probably more
\end{itemize}
\end{frame}

@@

@@

\begin{frame}{Linkage disequilibrium}
\begin{center}
\includegraphics[width=10cm]{ld.png}
\end{center}
\begin{itemize}
\item SNPs are correlated $\rightarrow$ tests are not independent
\item Bonferroni correction is overly stringent
\item What is the correlation structure in a 2D search?
\end{itemize}
\end{frame}

@@

\begin{frame}{A general formula for 2D thresholds?}
\begin{equation}
-\log_{10}(p_{T}) = a - \frac{a}{f_{1}(N)+ f_{2}(M)} \nonumber
\label{eqn:general_threshold}
\end{equation}
\begin{eqnarray}
a & = & -\log_{10} \left (\frac{0.05}{t^2/2} \right) = 12.68 \nonumber
\end{eqnarray}
where $t = 693138$ is the number of independent regions in the genome (Dudbridge \& Gusnanto 2008)
\end{frame}

@@

\begin{frame}{Very preliminary results}
\begin{itemize}
\item Is there much non-additive variance?
\item 71 probes have significant eQTLs after filtering for:
\begin{itemize}
\item Threshold of 16.5
\item Additive variance $\leq$ 60\% of total genetic variance
\item Relatively common SNPs (must have samples for all 9 genotype classes)
\item No LD between interacting SNPs
\end{itemize}
\item Number of significant epistatic pairs: \textbf{140}
\end{itemize}
\end{frame}

@@

{\tiny
PhD Funding:
\begin{itemize}
\item The Roslin Institute
\item University of Edinburgh
\item Biosciences KTN
\item Monsanto / Newsham Choice Genetics
\end{itemize}
PhD Supervisors:
\begin{itemize}
\item Chris Haley
\item Sara Knott
\item John Woolliams
\end{itemize}
Important people:
\begin{itemize}
\item Ricardo Pong-Wong
\item Athanasios Theocharidis
\item Wenhua Wei
\end{itemize}
Compute resources
\begin{itemize}
\item ECDF - {\tt eddie}
\item Daresbury Labs - {\tt cseht}
\item IVEC - {\tt fornax}
\item QBI cluster
\end{itemize}
}

@@

\begin{itemize}
\item Peter Visscher
\item Naomi Wray
\item Joseph Powell
\item Jian Yang
\item Allan Mcrae
\item Anita Goldinger
\item Hong Lee
\item Anna Vinkhuyzen
\item Guo-Bo Chen
\item Beben Benyamin
\item Zong Zhang
\item Enda Byrne
\item Marie-Jo Brion
\item Sven Stringer
\item Visit us at www.complextraitgenomics.com
\end{itemize}
}

>>>>

\documentclass{beamer}
%\usepackage{colortbl}
\usepackage[latin1]{inputenc}
\usepackage{comment}
%\usetheme{Warsaw}
\usetheme{Frankfurt}
\usecolortheme{dove}
\usepackage[absolute,overlay]{textpos} 
\usepackage{booktabs}
\usepackage{color}
\usepackage{threeparttable}
\usepackage{multirow}
%\usepackage[table]{xcolor}
%\definecolor{tableShade}{HTML}{F1F5FA}

\title[Epistasis in GWAS]{Genetic interactions in complex traits}
\author{Gibran Hemani}
\institute{The Roslin Institute, University of Edinburgh \\Diamantina Institute, University of Queensland \\Queensland Brain Institute}
\date{}
\begin{document}

@@

\section[Introduction]{Genome wide association studies}
\subsection{}

\begin{frame}{Missing heritability in GWAS}
\begin{center}
\begin{figure}
\includegraphics[width=7cm]{missing.jpg} \\
{\tiny Maher, 2008}
\end{figure}
\end{center}
\begin{itemize}
\item Many variants found and replicated from huge studies
\item Most have extremely small effect sizes
\item Very small proportion of estimated additive variance
\end{itemize}
\end{frame}

\section{Epistasis}
\subsection{}

@@

\begin{frame}{Examples}

\begin{figure}[htb]
 \centering
 \begin{minipage}[c]{0.38\textwidth}
  \centering
{\tiny
\begin{tabular}{cccc}
& AA & Aa & aa \\\hline
BB & $2\alpha + 2\beta$ & $\alpha + 2\beta$ & $2\beta$ \\
Bb & $2\alpha + \beta$ & $\alpha + \beta$ & $\beta$ \\
bb & $2\alpha$ & $\alpha$ & $0$ \\
&&&\\
\multicolumn{4}{l}{Additive + additive (not epistatic)}\\

\end{tabular} }
 \end{minipage}
 \begin{minipage}[c]{0.38\textwidth}
 \includegraphics[width=\textwidth]{additive.jpg}
 \end{minipage}
 \end{figure}

\begin{figure}[htb]
 \centering
 \begin{minipage}[c]{0.38\textwidth}
   \centering
{\tiny
\begin{tabular}{cccc}
& AA & Aa & aa \\\hline
BB & $\alpha$ & $\alpha$ & $\alpha$ \\
Bb & $\alpha$ & $\alpha$ & $\alpha$ \\
bb & $\alpha$ & $\alpha$ & $\beta$ \\
&&&\\
\multicolumn{4}{l}{Canalisation (epistatic)}\\
\end{tabular} }
 \end{minipage}
 \begin{minipage}[c]{0.38\textwidth}
 \includegraphics[width=\textwidth]{canalisation.jpg}
 \end{minipage}
\end{figure}
\end{frame}

\begin{frame}{Two dimensional GWAS}
\begin{center}
\includegraphics[width=7cm]{2dscan.png} \\
{\tiny Carlborg 2004}
\end{center}
\end{frame}

@@

\begin{frame}{Can epistasis maintain additive variance?}
\begin{itemize}
\item Hypothesis: Epistasis can maintain additive variance better than purely additive effects
\item How can maintained additive variance be detected most effectively?
\item What is the impact of low LD between causal variants and observed SNPs?
\end{itemize}
\end{frame}

\begin{frame}{Changes in variance under selection}
\begin{center}
\includegraphics[width=8cm]{sup_propadditive_det.png}
\end{center}
\end{frame}

\begin{frame}{Power to detect evolutionarily persistant variants}

\column{.7\textwidth}

\includegraphics[width=8cm]{sup_heritbars_sim.png}

\column{.3\textwidth}

When LD between causal variants and observed SNPs is high, 2D scan is most powerful for 46/54 non-additive patterns

\begin{frame}{Can epistasis maintain additive variance?}
\begin{itemize}
\item Hypothesis: Epistasis can maintain additive variance better than purely additive effects
\begin{itemize}
\item Yes, to a much greater extent
\end{itemize}
\item How can maintained additive variance be detected most effectively?
\begin{itemize}
\item 2 dimensional exhaustive searches
\end{itemize}
\item What is the impact of low LD between causal variants and observed SNPs?
\begin{itemize}
\item Upwards bias in reporting additive effects
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{The problem}
Number of interactions: \\
\begin{equation}
n(n-1) / 2 \nonumber
\end{equation}
500000 SNPs $\rightarrow$ 125 billion interactions \\
\begin{itemize}
\item PLINK performs 5000 tests per second $\rightarrow$ 3 years
\item FastEpistasis performs 36000 tests per second $\rightarrow$ 5 months
\item Optimised C code 125000 tests per second $\rightarrow$ 6 weeks
\item Parallelised on 8-core CPU $\rightarrow$ 5 days
\end{itemize}
\end{frame}

\begin{frame}{CPUs and hitting the `power wall'}
\begin{itemize}
\item All electrical power consumed eventually radiates as heat
\item Processor speed requires more power to improve
\begin{itemize}
\item Increased heat
\item Increased circuit resistance
\item Instability
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}{CPU limits}
\begin{center}
\includegraphics[height=6cm]{clockrates.png}
\end{center}
\end{frame}

\begin{frame}{Graphics Cards}
\begin{itemize}
\item Revenue of video game industry $>$ Hollywood film industry
\item Better games require better hardware
\begin{itemize}
\item Finer pixels rendered faster
\item Physics engines
\item Easily obtained by having many simple cores
\end{itemize}
\item Consumer level graphics cards have hundreds of very simple cores
\item Ideal for geometric parallelisation problems in science
\item Use OpenCL API to communicate with graphics cards

\end{itemize}
\end{frame}

\begin{frame}{Chip structures}
\includegraphics[height=2.5cm]{cpugpu.png}
\begin{itemize}
\item Many cores that can process data simultaneously
\item But all the cores are extremely simple
\begin{itemize}
\item Tiny L1 cache
\item Little communication between cores
\item Each core is relatively slow
\item Cores are divided into local groups, all of which must do the same thing
\end{itemize}

\end{itemize}
\end{frame}

@@

% mention epiGPU

@@

\begin{frame}{Multiple testing}
\begin{definition}
{\color{orange} Frequentist $p$-value: The probability of obtaining a test statistic at least as extreme as the one that was actually observed, assuming that the null hypothesis is true.}
\end{definition}
\begin{itemize}
\item \emph{e.g.} If 20 independent tests are performed then would expect one true null to have a $p$-value $\leq 0.05$.
\item So if significance with a false positive rate of $\alpha = 0.05$ is desired, then the threshold for the family of tests should be adjusted to $\alpha_{F} = 0.05 / 20 = 0.0025$.
\item 500000 SNPs being tested $\rightarrow \alpha_{F} = 0.05 / 500000 = 1 \times 10^{-7}$
\item Epistasis $\rightarrow \alpha_{F} = \frac{0.05}{n(n-1)/2} \approx 4 \times 10^{-13}$
\end{itemize}
\end{frame}

@@

\begin{frame}{Linkage disequilibrium}
\begin{center}
\includegraphics[width=10cm]{ld.png}
\end{center}
\begin{itemize}
\item SNPs are correlated $\rightarrow$ tests are not independent
\item Bonferroni correction is overly stringent
\end{itemize}
\end{frame}

@@

\begin{frame}{What is the correlation structure in a 2D search?}
\begin{center}
\includegraphics[width=7cm]{2dscan.png} \\
{\tiny Carlborg 2004}
\end{center}
\end{frame}

\begin{frame}{Epistatic permutations}
\begin{itemize}
\item How is the 2D threshold governed by:
\begin{itemize}
\item Number of SNPs being tested
\item Number of individuals in the study?
\end{itemize}
\item Difficult computational challenge
\begin{itemize}
\item Requires cluster of GPUs...
\item ...running for months!
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{A general formula for 2D thresholds?}
\begin{equation}
-\log_{10}(p_{T}) = a - a \left ( 
 s_{1} \exp \left ( 
  \frac {\ln(n) } { s_{2} } \right )
 + s_{3} \exp \left (
  \frac{ \ln(M)} {s_{4}} \right ) 
\right )^{-1} \nonumber
\label{eqn:general_threshold}
\end{equation}
\begin{eqnarray}
a & = & 12.68 \nonumber \\
s_{1} & = & 0.0258 \nonumber \\
s_{2} & = & 1.72 \nonumber \\
s_{3} & = & 0.0379 \nonumber \\
s_{4} & = & 2.33. \nonumber
\end{eqnarray}
\end{frame}

@@

\begin{frame}{Very preliminary results}
\begin{itemize}
\item Is there much non-additive variance?
\item 71 probes have significant eQTLs after filtering for:
\begin{itemize}
\item Threshold of 16.5
\item Additive variance $\leq$ 60\% of total genetic variance
\item Relatively common SNPs (must have samples for all 9 genotype classes)
\item No LD between interacting SNPs
\end{itemize}
\item Total of 140 significant independent loci
\end{itemize}
\end{frame}

@@

{\tiny
PhD Funding:
\begin{itemize}
\item The Roslin Institute
\item University of Edinburgh
\item Biosciences KTN
\item Monsanto / NCG
\end{itemize}
PhD Supervisors:
\begin{itemize}
\item Chris Haley
\item Sara Knott
\item John Woolliams
\end{itemize}
Important people:
\begin{itemize}
\item Ricardo Pong-Wong
\item Athanasios Theocharidis
\item Wenhua Wei
\end{itemize}
Compute resources
\begin{itemize}
\item ECDF - {\tt eddie}
\item Daresbury Labs - {\tt cseht}
\item IVEC - {\tt fornax}
\end{itemize}
}

@@

\begin{itemize}
\item Peter Visscher
\item Naomi Wray
\item \textbf{Joseph Powell}
\item Jian Yang
\item Allan Mcrae
\item Anita Goldinger
\item Hong Lee
\item Anna Vinkhuyzen
\item Guo-Bo Chen
\item Beben Benyamin
\item Zong Zhang
\item Enda Byrne
\item Marie-Jo Brion
\item Sven Stringer
\end{itemize}
}
